Fluent statistical computing interfaces for biological data analysis

### 2020-08-04 Rmd/01-intro.Rmd
reference fixes

The approach taken by the suite of software packages collectively known as the 
**tidyverse**  is an attempt to formalise aspects of the EDA process in the R
programming language under a single semantic known as *tidy data* 
[@baser; @tidyverse; @Wickham2014-jc]. Simply put, a *tidy data* set is a 
rectangular table where each row of the table corresponds to an observation, 
each column corresponds to a variable and each cell a value. There is a 
surprisingly large amount of utility that can be achieved with this definition. 
By having each column representing a variable, variables in the data can be 
mapped to graphical aesthetics of plots. This paradigm enables the grammar of graphics as implemented by **ggplot2** [@Wickham2016-gz; @Wilkinson2005-kq]. User interfaces as implemented by **tidyverse**, and in particular the **dplyr** package, are *fluent*;  they form a domain specific language (DSL) that gives users a mental model for performing and composing common data transformation tasks 
[@Wickham2017-dplyr; @FowlerFluent].

>>>>

The approach taken by the suite of software packages collectively known as the 
**tidyverse**  is an attempt to formalise aspects of the EDA process in the R
programming language under a single semantic known as *tidy data* 
[@r-core; @tidyverse; @Wickham2014-jc]. Simply put, a *tidy data* set is a 
rectangular table where each row of the table corresponds to an observation, 
each column corresponds to a variable and each cell a value. There is a 
surprisingly large amount of utility that can be achieved with this definition. 
By having each column representing a variable, variables in the data can be 
mapped to graphical aesthetics of plots. This paradigm enables the grammar of graphics as implemented by **ggplot2** [@Wickham2016-gz; @Wilkinson2005-kq]. User interfaces as implemented by **tidyverse**, and in particular the **dplyr** package, are *fluent*;  they form a domain specific language (DSL) that gives users a mental model for performing and composing common data transformation tasks 
[@Wickham2017-dplyr; @FowlerFluent].

### 2020-08-04 Rmd/03-fluentGenomics.Rmd
reference fixes

Throughout the workflow (Figure \@ref(fig:workflow)), we will use existing
Bioconductor infrastructure to understand these datasets. In particular, we
will emphasize the use of the Bioconductor packages **plyranges** and **tximeta**.
The **plyranges** package fluently transforms data tied to genomic ranges using
operations like shifting, window construction, overlap detection, etc. It is
described by @Lee2019 and leverages underlying core Bioconductor infrastructure
[@granges; @bioc] and the tidyverse design principles @tidyverse.

@@

The data used in this workflow is available from two packages: the **macrophage**
Bioconductor ExperimentData package and from the workflow package 
**fluentGenomics**.

@@

After we have read the `coldata.csv` file, we select relevant columns from this
table, create a new column called `files`, and transform the existing `line`
and `condition` columns into factors.  In the case of `condition`, we specify
the "naive" cell line as the reference level.  The `files` column points to the
quantifications for each observation -- these files have been gzipped, but
would typically not have the 'gz' ending if used from **Salmon** directly. One
other thing to note is the use of the pipe operator,`%>%`, which can be read as
"then", i.e. first read the data, *then* select columns, *then* mutate them.

Now we have a table summarizing the experimental design and the locations of
the quantifications. The following lines of code do a lot of work for the
analyst: importing the RNA-seq quantification (dropping *inferential replicates* 
in this case), locating the relevant reference transcriptome,
attaching the transcript ranges to the data, and fetching genome information.
Inferential replicates are especially useful for performing transcript-level
analysis, but here we will use a point estimate for the per-gene counts and
perform gene-level analysis.

The result is a _SummarizedExperiment_ object.

@@

On a machine with a working internet connection, the above command works
without any extra steps, as the **tximeta** function obtains any necessary
metadata via FTP, unless it is already cached locally. The **tximeta** package
can also be used without an internet connection, in this case the linked
transcriptome can be created directly from a *Salmon* index and gtf.

@@

One final note is that the _start_ of positive strand genes and the _end_ of
negative strand genes is now dictated by the genomic extent of the isoforms of
the gene (so the _start_ and _end_ of the reduced *GRanges*). Another
alternative would be to either operate on transcript abundance, and perform
differential analysis on transcript (and so avoid defining the TSS of a set of
isoforms), or to use gene-level summarized expression but to pick the most
representative TSS based on isoform expression.

@@

The ATAC-seq counts have already been normalized with **cqn** [@Hansen2012] and
$log_2$ transformed. Loading the **cqn**-normalized matrix of $log_2$ transformed read
counts takes ~30 seconds and loads an object of ~370 Mb. We set the column
names so that the first column contains the rownames of the matrix, and the
remaining columns are the sample identities from the `atac_coldata` object.

@@

We now output the results as a _GRanges_ object, and due to the conventions of
**plyranges**, we construct a new column called _gene\_id_ from the row names of
the results. Each row now contains the genomic region (_seqnames_, _start_,
_end_, _strand_) along with corresponding metadata columns (the _gene_id_ and
the results of the test). Note that **tximeta** has correctly identified the
reference genome as "hg38", and this has also been added to the **GRanges** along
the results columns. This kind of book-keeping is vital once overlap operations
are performed to ensure that **plyranges** is not comparing across incompatible
genomes.

@@

From this, we can restrict the results to those that meet our FDR threshold and
select (and rename) the metadata columns we're interested in:

@@

The following section describes the process we have used for generating a
_GRanges_ object of differential peaks from the ATAC-seq data in @alasoo. The 
code chunks for the remainder of this section are not run.

@@

Now we would like to modify our gene ranges so they contain the 10 kilobases on
either side of their transcription start site (TSS). There are many ways one
could do this, but we prefer an approach via the anchoring methods in
**plyranges**. Because there is a mutual dependence between the start, end,
width, and strand of a *GRanges* object, we define anchors to fix one of
_start_ and _end_, while modifying the _width_. As an example, to extract just
the TSS, we can anchor by the 5' end of the range and modify the width of the
range to equal 1.

@@

To combine the DE genes with the DA peaks, we perform a left overlap join. This
returns to us the `all_genes` ranges (potentially with duplication), but with
the metadata columns from those overlapping DA peaks.  For any gene that has no
overlaps, the DA peak columns will have `NA`'s.

@@

For each gene (that is the combination of chromosome, the start, end, and
strand), and the "origin" (DE vs not-DE) we can compute the distinct number of
peaks for each gene and the maximum peak based on LFC. This is achieved via
`reduce_ranges_directed()`, which allows an aggregation to result in a _GRanges_
object via merging neighboring genomic regions. The use of the directed suffix
indicates we're maintaining strand information. In this case, we are simply
merging ranges (genes) via the groups we mentioned above. We also have to
account for the number of resamples we have performed when counting if there
are any peaks, to ensure we do not double count the same peak:

@@

For all variables except for the _origin_ column we divide the first row's
values by the second row, which will be the enrichment of peaks in DE genes
compared to other genes. This requires us to reshape the summary table from
long form back to wide form using the **tidyr** package. First we pivot the
results of the _peak\_count_ columns into name-value pairs, then pivot again to
place values into the _origin_ column. Then we create a new column with the
relative enrichment:

>>>>

Throughout the workflow (Figure \@ref(fig:workflow)), we will use existing
Bioconductor infrastructure to understand these datasets. In particular, we
will emphasize the use of the Bioconductor packages **plyranges** and **tximeta**.
The **plyranges** package fluently transforms data tied to genomic ranges using
operations like shifting, window construction, overlap detection, etc. It is
described by @Lee2019 and leverages underlying core Bioconductor infrastructure
[@granges; @bioc] and the **tidyverse** design principles @tidyverse.

@@

The data used in this workflow is available from two packages: the **macrophage**
Bioconductor **ExperimentData** package and from the workflow package 
**fluentGenomics**.

@@

After we have read the `coldata.csv` file, we select relevant columns from this
table, create a new column called `files`, and transform the existing `line`
and `condition` columns into factors.  In the case of `condition`, we specify
the "naive" cell line as the reference level.  The `files` column points to the
quantifications for each observation -- these files have been gzipped, but
would typically not have the 'gz' ending if used from **Salmon** directly. One
other thing to note is the use of the pipe operator, `%>%`, which can be read as
"then", i.e. first read the data, *then* select columns, *then* mutate them.

Now we have a table summarizing the experimental design and the locations of
the quantifications. The following lines of code do a lot of work for the
analyst: importing the RNA-seq quantification (dropping *inferential replicates* 
in this case), locating the relevant reference transcriptome,
attaching the transcript ranges to the data, and fetching genome information.
Inferential replicates are especially useful for performing transcript-level
analysis, but here we will use a point estimate for the per-gene counts and
perform gene-level analysis. The result is a _SummarizedExperiment_ object.

@@

On a machine with a working internet connection, the above command works
without any extra steps, as the **tximeta** function obtains any necessary
metadata via FTP, unless it is already cached locally. The **tximeta** package
can also be used without an internet connection, in this case the linked
transcriptome can be created directly from a **Salmon** index and gtf.

@@

One final note is that the _start_ of positive strand genes and the _end_ of
negative strand genes is now dictated by the genomic extent of the isoforms of
the gene (so the _start_ and _end_ of the reduced *GRanges*). Another
alternative would be to either operate on transcript abundance, and perform
differential analysis on transcript, and so avoid defining the transcription start
site (TSS) of a set of isoforms, or to use gene-level summarized expression 
but to pick the most representative TSS based on isoform expression.

@@

The ATAC-seq counts have already been normalized with **cqn** [@Hansen2012] and
$\log_2$ transformed. Loading the **cqn**-normalized matrix of $\log_2$ transformed read
counts takes ~30 seconds and loads an object of ~370 Mb. We set the column
names so that the first column contains the rownames of the matrix, and the
remaining columns are the sample identities from the `atac_coldata` object.

@@

We now output the results as a _GRanges_ object, and due to the conventions of
**plyranges**, we construct a new column called _gene\_id_ from the row names of
the results. Each row now contains the genomic region ( _seqnames_, _start_,
_end_, _strand_) along with corresponding metadata columns (the _gene_id_ and
the results of the test). Note that **tximeta** has correctly identified the
reference genome as "hg38", and this has also been added to the **GRanges** along
the results columns. This kind of book-keeping is vital once overlap operations
are performed to ensure that **plyranges** is not comparing across incompatible
genomes.

@@

From this, we can restrict the results to those that meet our FDR threshold and
select (and rename) the metadata columns we are interested in:

@@

The following section describes the process we have used for generating a
_GRanges_ object of differential peaks from the ATAC-seq data in @alasoo. The 
code chunks for the remainder of this section are optional.

@@

Now we would like to modify our gene ranges so they contain the 10 kilobases on
either side of their TSS. There are many ways one
could do this, but we prefer an approach via the anchoring methods in
**plyranges**. Because there is a mutual dependence between the start, end,
width, and strand of a *GRanges* object, we define anchors to fix one of
_start_ and _end_, while modifying the _width_. As an example, to extract just
the TSS, we can anchor by the 5' end of the range and modify the width of the
range to equal 1.

@@

To combine the DE genes with the DA peaks, we perform a left overlap join. This
returns to us the `all_genes` ranges (potentially with duplication), but with
the metadata columns from those overlapping DA peaks.  For any gene that has no
overlaps, the DA peak columns will have `NA` values.

@@

For each gene (that is the combination of chromosome, the start, end, and
strand), and the "origin" (DE vs not-DE) we can compute the distinct number of
peaks for each gene and the maximum peak based on LFC. This is achieved via
`reduce_ranges_directed()`, which allows an aggregation to result in a _GRanges_
object via merging neighboring genomic regions. The use of the directed suffix
indicates we are maintaining strand information. In this case, we are simply
merging ranges (genes) via the groups we mentioned above. We also have to
account for the number of resamples we have performed when counting if there
are any peaks, to ensure we do not double count the same peak:

@@

For all variables except for the _origin_ column we divide the first rows
values by the second row, which will be the enrichment of peaks in DE genes
compared to other genes. This requires us to reshape the summary table from
long form back to wide form using the **tidyr** package. First we pivot the
results of the _peak\_count_ columns into name-value pairs, then pivot again to
place values into the _origin_ column. Then we create a new column with the
relative enrichment:

### 2020-08-04 Rmd/04-intron.Rmd
reference fixes

Although there are many flexible and powerful software tools, like BEDtools or deepTools, for exploring and estimating coverage from common genomic data formats [@Quinlan2010-gc; @Ramirez2014-sh], it is advantageous to have tooling
that is tightly coupled to a statistical computing language such as R. This allows interoperability between other software packages for data wrangling, visualisation and modelling within the ecosystem of the language that may not
be possible with a single command line tool. There are also gains in reproducibility as analysts do not have to move between multiple software suites to explore their processed transcriptomics data.

@@

(ref:si-fig) An overview of the **superintronic** and **plyranges** workflow. Coverage is estimated directly using the design matrix that contains a source column pointing to the locations of BAM files. The long-form representation is output as a **GRanges** object, and contains columns that were part of the design. Additional annotations are add with join functions, here we show the particular case of expanding the coverage **GRanges** to include exonic and intronic parts of a gene. This object can be further analysed using **plyranges** and our data descriptors approach, and then descriptors can be visualised as a scatter plot matrix with the `GGally` and `ggplot2` packages [@ggally; @Wickham2016-gz]. Coverage traces can be directly generated with **superintronic** and collapsed over parts of the design matrix to identify differences between groups.

@@

Our zebrafish workflow shows that our approach using **superintronic** 
and **plyranges** is able to uncover interesting biological signals in a
purely data-driven manner. We did not include additional information
that could have been useful when deriving our selected genes, such as
sequence motifs for U12 class of introns, or exploit the experimental
design to find differential IR like profiles. However, if that was of interest,
one could look at the overlaps, like we did in figure \@ref(fig:si-upset),
or combine our data descriptors with external estimates using **limma** 
[@Ritchie2015-mj], like our proposed index method in @Lee2019-mf. The gene
candidates obtain by our thresholds have been validated by the Heath lab
using qPCR.

>>>>

Although there are many flexible and powerful software tools, like **BEDtools** or **deepTools**, for exploring and estimating coverage from common genomic data formats [@Quinlan2010-gc; @Ramirez2014-sh], it is advantageous to have tooling
that is tightly coupled to a statistical computing language such as R. This allows interoperability between other software packages for data wrangling, visualisation and modelling within the ecosystem of the language that may not
be possible with a single command line tool. There are also gains in reproducibility as analysts do not have to move between multiple software suites to explore their processed transcriptomics data.

@@

(ref:si-fig) An overview of the **superintronic** and **plyranges** workflow. Coverage is estimated directly using the design matrix that contains a source column pointing to the locations of BAM files. The long-form representation is output as a _GRanges_ object, and contains columns that were part of the design. Additional annotations are added with join functions, here we show the particular case of expanding the coverage _GRanges_ to include exonic and intronic parts of a gene. This object can be further analysed using **plyranges** and our data descriptors approach, and then descriptors can be visualised as a scatter plot matrix with the **GGally** and **ggplot2** packages [@ggally; @Wickham2016-gz]. Coverage traces can be directly generated with **superintronic** and collapsed over parts of the design matrix to identify differences between groups.

@@

Our zebrafish workflow shows that our approach using **superintronic** 
and **plyranges** is able to uncover interesting biological signals in a
purely data-driven manner. We did not include additional information
that could have been useful when deriving our selected genes, such as
sequence motifs for U12 class of introns, or exploit the experimental
design to find differential IR like profiles. However, if that was of interest,
one could look at the overlaps, like we did in figure \@ref(fig:si-upset),
or combine our data descriptors with external estimates using **limma** 
[@Ritchie2015-mj], like our proposed index method in @Lee2019-mf. The gene
candidates obtained by our thresholds have been validated by the Heath lab
using qPCR.
